---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 3 - Due date 02/08/22"
author: "Rob Kravec"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,
                      tidy.opts=list(width.cutoff=80), tidy=FALSE,
                      warning = F, message = F) 
```

## Questions 

Consider the same data you used for A2. 
The data comes from the US Energy Information and Administration and 
corresponds to the January 2022 **Monthly** Energy Review. Once again you will 
work only with the following columns: Total Biomass Energy Production, 
Total Renewable Energy Production, Hydroelectric Power Consumption. 
Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". 
Install these packages, if you haven't done yet. Do not forget to load them 
before running your script, since they are NOT default packages.

```{r load-packages}
library(tidyverse)
library(forecast)
library(Kendall)
library(tseries)
library(readxl)
library(patchwork)
```

First, I'll create the time series object requested, pulling code from A02.

```{r create-ts}
# Read in data
file_path = paste0('../Data/Table_10.1_Renewable_Energy_Production_and', 
                   '_Consumption_by_Source.xlsx')
data <- read_excel(path = file_path, sheet = "Monthly Data", skip = 10, 
                   na = "Not Available")

# Remove first row, which contains units for each column
data <- data[-1, ]

# Rename relevant columns
data <- data %>% 
  rename(Biomass_prod = 'Total Biomass Energy Production',
         Renewable_prod = 'Total Renewable Energy Production',
         Hydro_consumption = 'Hydroelectric Power Consumption')

# Select columns
data_small <- data %>% 
  select(Biomass_prod, Renewable_prod,
         Hydro_consumption) 

# Convert data types to numeric
data_small <- sapply(data_small, as.numeric) %>% 
  as_tibble()

# Create df for plotting
plot_df <- cbind(Month = data$Month, data_small)

# Create ts object
data_ts <- ts(data = data_small, start = c(1973, 1), end = c(2021, 9), 
              frequency = 12)

# Show first 6 rows of ts object
head(data_ts)
```


## Trend Component

### Q1

Create a plot window that has one row and three columns. And then for each 
object on your data frame, fill the plot window with time series plot, ACF 
and PACF. You may use the some code form A2, but I want all three plots on the 
same window this time. (Hint: use par() function)

```{r q1}
# Define functions to create desired plot panel
plot_acf <- function(ts, lag_amt, title = "ACF") {
  # Prepare data
  acf_data <- data.frame(lag = 1:lag_amt, 
                       acf = Acf(ts, lag.max = lag_amt, 
                                 plot = F)$acf[2:(lag_amt + 1)])
  
  # Create plot
  acf_plt <- ggplot(data = acf_data, mapping = aes(x = lag, y = acf)) + 
    geom_bar(stat = 'identity') +
    labs(x = 'Lag', y = '', title = title) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.title.y = element_blank())
  
  # Return plot
  return(acf_plt)
}

plot_pacf <- function(ts, lag_amt, title = "PACF") {
  # Prepare data
  pacf_data <- data.frame(lag = 1:lag_amt, 
                       pacf = pacf(ts, 
                                 plot = F, 
                                 lag.max = lag_amt)$acf)
  
  # Create plot
  pacf_plt <- ggplot(data = pacf_data, mapping = aes(x = lag, y = pacf)) + 
    geom_bar(stat = 'identity') +
    labs(x = 'Lag', y = '', title = title) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.title.y = element_blank()) 
  
  # Return plot
  return(pacf_plt)
}


plt_three <- function(plot_df, col_num, lag_amt, ts_y_lab) {
  # Designate column of interest for time series plot
  plot_df <- plot_df %>% 
    mutate(ts_plot = plot_df[, col_num])
  
  # Create time series plot
  ts_plt <- ggplot(data = plot_df, mapping = aes(x = Month, y = ts_plot)) +
    geom_line() +
    labs(x = 'Date', y = ts_y_lab, 
         title = 'Time Series Plot') +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5))
  
  # Create ACF plot
  acf_plt <- plot_acf(plot_df[, col_num], lag_amt)
  
  # Create PACF plot
  pacf_plt <- plot_pacf(plot_df[, col_num], lag_amt) 
  
  # Return all 3 plots in a single row using patchwork syntax
  return(ts_plt + acf_plt + pacf_plt)
}
```

```{r q1-ii, fig.height = 3}
# Create requested plot panels
plt_three(plot_df, 2, 40, 'Biomass production [Trillion BTU]')
plt_three(plot_df, 3, 40, 'Renewable production [Trillion BTU]')
plt_three(plot_df, 4, 40, 'Hydropower consumption [Trillion BTU]')
```

### Q2

From the plot in Q1, do the series Total Biomass Energy Production, 
Total Renewable Energy Production, Hydroelectric Power Consumption appear to 
have a trend? If yes, what kind of trend?

`Biomass production` and `Renewable production` appear to have a strong positive
trend, while `Hydropower consumption` appears to have a weak negative trend. 
From the results in `Q4`, we can describe these trends as deterministic
trends, since they can all be modeled effectively by a single straight line.
By "effectively," I mean that the detrended series have approximately 
horizontal trend lines near y = 0.

### Q3

Use the *lm()* function to fit a linear trend to the three time series. 
Ask R to print the summary of the regression. Interpret the regression 
output, i.e., slope and intercept. Save the regression coefficients for 
further analysis.

```{r q3}
# Define time vector
t <- 1:nrow(data)

# Initialize dataframe with output from regressions
df_q3 <- data.frame(Intercept = c(0, 0, 0),
                    Slope = c(0, 0, 0),
                    row.names = c("Biomass", "Renewables", "Hydro"))

# Perform regressions and print summaries, as instructed
for (i in 1:3) {
  model <- lm(data_ts[, i] ~ t)
  print(colnames(data_ts)[i])
  print(summary(model))
  df_q3[i, 1] <- summary(model)$coefficients[1]
  df_q3[i, 2] <- summary(model)$coefficients[2]
}

# Display results in tabular format
df_q3
```

Regression interpretation:

- **Biomass**: Starts at `r round(df_q3[1, 1], 3)` Trillion BTU in Jan 1973 and 
increases
by `r round(df_q3[1, 2], 3)` Trillion BTU each month
- **Renewables**: Starts at `r round(df_q3[2, 1], 3)` Trillion BTU in Jan 1973 
and increases
by `r round(df_q3[2, 2], 3)` Trillion BTU each month
- **Hydropower**: Starts at `r round(df_q3[3, 1], 3)` Trillion BTU in Jan 1973 
and decreases
by `r -round(df_q3[3, 2], 3)` Trillion BTU each month

### Q4

Use the regression coefficients from Q3 to detrend the series. Plot the 
detrended series and compare with the plots from Q1. What happened? 
Did anything change?

Below, I plot the original time series and detrended time series (both with 
trend lines) for the 3 variables of interest. For each we see evidence of 
deterministic trends, given that the detrended series have approximately
horizontal trend lines near y = 0. 

There still do appear to be some patterns in the detrended data (e.g., cycles over
long time horizons for `biomass production`, seasonal variation for
`hydropower consumption`), but there is at least no longer a clear up/downward
trend in each series

```{r q4, fig.height = 2.5}
# Generate plots in a loop
for (i in 1:3) {
  
  # Detrend series (and save results for next exercise)
  detrended <- data_ts[, i] - (t * df_q3[i, 2] + df_q3[i, 1])
  detrended_name <- paste0("detrended", i)
  assign(detrended_name, detrended)
  
  # Create dataframe for plotting
  plot_df_4 <- data.frame(Month = data$Month,
                          Original = data_ts[, i] %>% as.numeric(),
                          Detrended = detrended %>% as.numeric()) 
  
  # Generate plot
  plt <- ggplot(data = plot_df_4, mapping = aes(x = Month)) +
    geom_line(mapping = aes(y = Original, color = "Original")) +
    geom_smooth(mapping = aes(y = Original), method = 'lm', color = 'black',
                se = F, linetype = 2) +
    geom_line(mapping = aes(y = Detrended, color = "Detrended")) +
    geom_smooth(mapping = aes(y = Detrended), method = 'lm', color = 'black',
                se = F, linetype = 2) +
    labs(x = "Date", 
         title = paste0(colnames(data_small[, i]), " [Trillion BTU]"), 
         color = "Series") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank())
  
  # Display plot
  print(plt)
}
```


### Q5

Plot ACF and PACF for the detrended series and compare with the plots from 
Q1. Did the plots change? How?

For this question, I create a 2x2 grid of plots for each variable of interest,
where the top row of each grid shows the original ACF and PACF plots, while 
the bottom row shows the detrended ACF and PACF plots.

Some observations:

- For all three variables, detrended ACF values appear to be slightly lower than 
original ACF values
- It is more challenging to diagnose a consistent difference in PACF (aside
from lag 1). I can at least say that the PACF plots
look different after detrending and that I'm unable to pick up on a 
pattern
- Seasonal components appear more pronounced after detrending. This effect
is particularly clear for the detrended `renewables` plot but is present for 
all three variables

```{r q5, fig.height = 4}
# Create lists that enable plotting in a loop
dfs_5 <- list(detrended1, detrended2, detrended3)
name_list <- c("Biomass", "Renewables", "Hydro")

# Make plots
for (i in 1:3) {
  print((plot_acf(data_ts[, i], 40, paste0("Original ACF - ", name_list[i])) +
     plot_pacf(data_ts[, i], 40, paste0("Original PACF - ", name_list[i]))) /
    (plot_acf(dfs_5[[i]], 40, paste0("Detrended ACF - ", name_list[i])) + 
       plot_pacf(dfs_5[[1]], 40, paste0("Detrended PACF - ", name_list[i])))) 
}
```

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 
to answer Q6 to Q8.

### Q6

Do the series seem to have a seasonal trend? Which series? Use function 
*lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to 
this/these time series. Ask R to print the summary of the regression. Interpret 
the regression output. Save the regression coefficients for further analysis.

Based on the original ACF plots, it's clear that `Hydropower` contains a 
seasonal trend. I originally wasn't sure whether `Renewables` contained a 
seasonal trend, so I fit a seasonal means model (not shown), and none of the
coefficients had a p-value less than 0.05, indicating an absence of seasonal
trend. I then repeated the exercise for `Biomass`, really just for fun, and
found the same result. Therefore, the output and plots shown below pertain
to the `Hydropower` series.

Additionally, I noted in `Q5` that the detrended series appeared to have more
accentuated seasonal trends than the original series. I generated seasonal 
means models for detrended `Renewables` and `Biomass` series (also not shown),
and some of the coefficients had significant p-values.

Interpreting the `Hydropower` regression output, the `Intercept` term 
corresponds to the 
mean value for December, and all other coefficients are adjustments from 
that baseline. Using a p-value threshold of $\alpha = 0.05$, we observe 
hydropower consumption significantly above the December baseline in January
and Mar-Jun. We also observe hydropower consumption significantly below the 
December baseline in Aug-Nov. These results indicate the likely presence of
a seasonal trend.

```{r q6}
# Fit model, and display summary
i <- 3
dummies <- seasonaldummy(data_ts[, i])
model <- lm(data_ts[, i] ~ dummies)
summary(model)

# Save regression coefficients
q6_coef <- summary(model)$coefficients[,1]
```

### Q7

Use the regression coefficients from Q6 to deseason the series. Plot the 
deseason series and compare with the plots from part Q1. Did anything change?

The deseasoned `hydropower` series definitely has less of the predictable 
choppiness that characterizes the seasonal trend in the original series. The
overall shape (e.g., sustained higher levels 1995-2000) and negative trend of 
the original series is maintained in the deseasoned version.

```{r q7, fig.height = 2.5}
# Deseason hydropower consumption time series
seasonal_component <- dummies %*% q6_coef[2:12] + q6_coef[1]
deseasoned <- data_small[, 3] - seasonal_component
colnames(deseasoned) <- "Deseasoned"

# Create dataframe for plotting
plot_df_7 <- cbind(plot_df, deseasoned)

# Plot time series on same plot
ggplot(data = plot_df_7, mapping = aes(x = Month)) +
    geom_line(mapping = aes(y = Hydro_consumption, color = "Original")) +
    geom_smooth(mapping = aes(y = Hydro_consumption), 
                method = 'lm', color = 'black',
                se = F, linetype = 2) +
    geom_line(mapping = aes(y = Deseasoned, color = "Deseasoned")) +
    geom_smooth(mapping = aes(y = Deseasoned), method = 'lm', color = 'black',
                se = F, linetype = 2) +
    labs(x = "Date", 
         title = "Hydropower Consumption [Trillion BTU]", 
         color = "Series") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank())
```

### Q8

Plot ACF and PACF for the deseason series and compare with the plots from Q1. 
Did the plots change? How?

Yes, the plots changed a lot! The wave-like appearance in the original ACF plot
is essentially gone in the deseasoned ACF plot. The deseasoned series
has lower PACF values overall without any discernable wave-like pattern, 
suggesting that the vast majority of partial autocorrelation is coming from 
lag one.

```{r q8, fig.height = 4}
(plot_acf(plot_df_7[, 4], 40, "Original ACF - Hydro") +
     plot_pacf(plot_df_7[, 4], 40, "Original PACF - Hydro")) /
    (plot_acf(plot_df_7[, 5], 40, "Deseasoned ACF - Hydro") + 
       plot_pacf(plot_df_7[, 5], 40, "Deseasoned PACF - Hydro"))
```



