---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 2 - Due date 01/26/22"
author: "Rob Kravec"
output: pdf_document
geometry: margin=2.54cm
---

## R packages

```{r load-packages, message = F, warning = F}
library(tidyverse)
library(forecast)
library(tseries)
library(readxl)
library(lubridate)
```

## Data set information  

I'll read in the data using the `read_excel` function from the `readxl` 
package because this function allows me to specify the sheet that I want to 
read from the `.xlsx` file. Though not specified in the assignment, I'll use
the `Monthly Data`. The amount of data is still pretty small, and I can always
aggregate to the annual view, if needed.

```{r import-data}
# Read in data
file_path = paste0('../Data/Table_10.1_Renewable_Energy_Production_and', 
                   '_Consumption_by_Source.xlsx')
data <- read_excel(path = file_path, sheet = "Monthly Data", skip = 10, 
                   na = "Not Available")

# Remove first row, which contains units for each column
# Remove second row, which does not specify a date for the data (though Jan 
# 1973 seems likely)
data <- data[-1, ]
```


## Question 1

You will work only with the following columns: Total Biomass Energy Production, 
Total Renewable Energy Production, Hydroelectric Power Consumption. Create a 
data frame structure with these three time series only. Use the command head() 
to verify your data.

```{r question-1}
# Select columns
data_q1 <- data %>% 
  select('Total Biomass Energy Production', 'Total Renewable Energy Production',
         'Hydroelectric Power Consumption') 

# Make column names shorter and remove spaces
data_q1 <- data_q1 %>% 
  rename(Biomass_prod = 'Total Biomass Energy Production',
         Renewable_prod = 'Total Renewable Energy Production',
         Hydro_consumption = 'Hydroelectric Power Consumption')

# Convert data types to numeric
data_q1 <- sapply(data_q1, as.numeric) %>% 
  as_tibble()

# Show first six rows
head(data_q1)
```


## Question 2

Transform your data frame in a time series object and specify the starting 
point and frequency of the time series using the function ts().

```{r question-2}
# Frequency of 12 chosen to represent monthly data
data_ts <- ts(data = data_q1, start = c(1973, 1), end = c(2021, 9), 
              frequency = 12)

# Display result
head(data_ts)
```


## Question 3

Compute mean and standard deviation for these three series.

```{r question-3}
# Define function that returns mean and standard deviation
mean_sd <- function(x) {
  c(mean(x), sd(x))
}

# Calculate mean and standard deviation for each column
mean_sd_results <- sapply(data_ts, mean_sd)

# Rename rows
row.names(mean_sd_results) <- c('Mean', 'Standard_deviation')

# Display results
mean_sd_results
```


## Question 4

Display and interpret the time series plot for each of these variables.

For this question, I decided to make a faceted plot and denote the mean value
of each time series with a dashed black line. A few observations are
apparent:

- While Total `Renewable Energy Production` and `Total Biomass Energy Production`
rose substantially from 1973-2021, `Hydroelectric Power Consumption` decreased
- Given the rise of `Total Biomass Energy Production` and the stagnation of
`Hydroelectric Power Consumption`, the mean biomass production is greater than
the mean hydroelectric consumption over the given time period. The mean value
of `Total Renewable Energy Production` is obviously the highest of the three 
categories shown, given that the biomass and hydroelectric variables are two
of the inputs of the sum that produces `Total Renewable Energy Production`
- The gap between `Renewable Energy Production` and 
`Total Biomass Energy Production` + `Hydroelectric Power Consumption` grew
over time, denoting the rise of other renewable energy technologies (e.g., 
wind, solar)

```{r question-4}
# Create dataframe with means
mean_df <- data.frame(name = c('Biomass', 'Renewables', 'Hydro'), 
                      means = mean_sd_results[1,])

# Generate requested plot
data %>% 
  select(Month, 'Total Biomass Energy Production',
         'Total Renewable Energy Production',
         'Hydroelectric Power Consumption') %>% 
  rename(Biomass = 'Total Biomass Energy Production',
         Renewables = 'Total Renewable Energy Production',
         Hydro = 'Hydroelectric Power Consumption') %>% 
  pivot_longer(cols = !Month) %>%
  mutate(Month = ymd(Month), value = as.numeric(value)) %>% 
  ggplot(mapping = aes(x = Month, y = value, color = name)) +
  geom_line() +
  facet_grid(name~.) +
  geom_hline(data = mean_df, mapping = aes(yintercept = means),
             color = 'black', linetype = 2) +
  labs(x = 'Time', y = 'Production / Consumption [Trillion BTU]', 
       title = 'Biomass and Renewable Production, Hydro Consumption') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "None") 
```


## Question 5

Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r question-5}

```


## Question 6

Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r question-6}

```


## Question 7

Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r question-7}

```


